---
title       : Visualizing moving averages
subtitle    : Making sense of a stream discharge time series
author      : J. M. Branlund
job         : 
framework   : io2012 # {landslide, flowtime, io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## Stream discharge

```{r dischargeplot, echo=FALSE, fig.width=8, fig.height=4}
streamFlow <- read.csv("05543500discharge", skip=28, header=FALSE, sep="\t")
streamFlow <- subset(streamFlow, V5 == "A")
streamFlow <- streamFlow[,1:4]
colnames(streamFlow) <- c("agency", "site", "date", "discharge")
streamFlow$date <- as.Date(streamFlow$date)
xmin <- as.Date("1950-01-01")
xmax <- as.Date("1990-01-01")
streamFlow <- subset(streamFlow, date > xmin)
streamFlow <- subset(streamFlow, date < xmax)
plot(streamFlow$date, streamFlow$discharge, type="l", col=grey(.5), ylab="Discharge (cubic feet per sec)", xlab="Year", xaxt="n")
axis.Date(1, at = seq(xmin, xmax, 365))
```

- the volume of water that flows through in a given time
- shown above for USGS gage station 05543500 (Illinois River)
- measured continually at many sites by the USGS (United States Geological Survey). The average daily discharges are available online at http://waterdata.usgs.gov/nwis.

--- .class #id 

## The problem

Ideally, one could answer questions using these time series, like:  
- "How did stream flow change due to levee construction?"
- "Are seasonal precipitation patterns responsible for streamflow?"
- "Has discharge changed in response to global warming?"


<font color = blue><b>BUT... </b></font> Patterns in streamflow with time are hard to discern from the time-series plot.

---

## A solution - the moving average

- Calculating a moving average is one way to tease out patterns in the data. 

- For a given date, the moving average is the average of values from that date and <em>n</em> previous dates (more data are averaged when <em>n</em> is larger). In R, this is done with the <em>filter</em> function. 

- Including different numbers of days may show different patterns, for example, changes in discharge due to seasonal precipitation patterns (a smaller <em>n</em> value) versus global climate change (larger <em>n</em>). 

---

## Importance of <em>n</em>

Below, the top graph shows a moving average calculated with <font color="blue">60 days (~2 months) </font>of data, whereas the bottom graph uses <font color="blue">360 days (~1 year)</font> of data. (Discharge is in cubic feet per second.)

```{r smallN, echo=FALSE, fig.width=8, fig.height=5}
layout(matrix(c(1,1,2,2), 2, 2, byrow = TRUE))
plot(streamFlow$date, streamFlow$discharge, type="l", col=grey(.5), ylab="Discharge", ylim = c(0, 4e4), xlab="", xaxt="n")
axis.Date(1, at = seq(xmin, xmax, 365))
Phil <- rep(1/60, 60)
AvgDischarge <- filter(streamFlow$discharge, Phil, sides=1)
lines(streamFlow$date, AvgDischarge, col="red", lwd=2)
plot(streamFlow$date, streamFlow$discharge, type="l", col=grey(.5), ylab="Discharge", ylim = c(0, 3e4), xlab="Year", xaxt="n")
axis.Date(1, at = seq(xmin, xmax, 365))
Phil <- rep(1/360, 360)
AvgDischarge <- filter(streamFlow$discharge, Phil, sides=1)
lines(streamFlow$date, AvgDischarge, col="red", lwd=2)
```

---

## An app to explore moving averages

The app, <font color="blue"><em>Using moving averages to find patterns in stream discharge</em></font>, allows the user to:

- change the value of <em>n</em> (number of days used in the moving average), and compare the resulting moving average with the data (which can be hidden for better viewing)
- change the range of dates and maximum discharge graphed, to better see the plotted data
- view lists of high discharge in the selected time range (in the Summary tab)

Future iterations of the app will:
- Allow users to select data from different stream gages
- Show a table of the peaks visible in the moving average graph

